### 2013-03-24 doc/sosp13/relatedwork.tex
minor edits to related work

\paragraph{Distributed graph processing} While \Grappa is a general runtime
system for any large-scale concurrent application, one of its \checkme{killer
applications is graph analysis}. Therefore, distributed graph processing
frameworks like Pregel~\cite{pregel:2010} and Distributed
GraphLab~\cite{distgraphlab:vldb12} are related. Pregel adopts a
bulk-synchronous parallel (BSP) execution model, which makes it inefficient on
workloads that could prioritize vertices. GraphLab overcomes this limitation with an execution mode
that schedules vertex computations individually, allowing prioritization, which
gives faster convergence in a variety of iterative algorithms. GraphLab,
however, imposes a rigid computation model where programmers must express
computation as transformations on a vertex and its edge list only, with
information only from adjacent vertexes. Pregel is only slightly less
restrictive, as the input data can be any vertex in the graph. Grappa also
supports dynamic parallelism with asynchronous execution, but parallelism is
expressed as tasks or loop iterations, which is a far more general programming
model for irregular computation tasks. PowerGraph~\cite{powergraph:osdi12} improves
the performance of GraphLab for real-world graphs with power-law degree distributions by
using a vertex cut for graph partitioning. If the size of the partial result of 
applying a subset of edge data to the vertex is constant, then there is a good bound on the communication
required to keep vertex replicas coherent. We do not explore these algorithmic techniques
in this work, but they are complimentary and would improve the performance of graph applications on Grappa.

\paragraph{Global memory} Grappa supports a global memory namespace, presenting a programming abstraction similar to that of previous software distributed shared memory (DSM) systems but with differences attuned to irregular application demands.

Many traditional software DSM systems are page based~\cite{Treadmarks,munin}
and aim to hide the fact that they are built in software from applications by
exploiting the processor's paging mechanisms, therefore relying heavily on
locality. Instead, Grappa, like partitioned global address space (PGAS)
models, implements its DSM at the language, rather than the system level.
Languages such as Chapel~\cite{Chamberlain:2007}, X10~\cite{X10:2005}, and
UPC~\cite{upc:2005} make accesses to shared structures look like normal
references. As we describe later, Grappa chooses a middle ground, where global
addresses are explicit in the API and local accesses are emitted
conventionally by the compiler.

Past DSM work was mostly page-based, so it could take advantage of a network's RDMA support to move large page-sized
blocks of data from node to node \TODO{citation}. However, when commodity networks move small
blocks of data (a few bytes), only a fraction of the available bandwidth is
achieved. \TODO{probably want to reframe this discussion in view of Grappa using RDMA differently}

In summary, while Grappa's DSM system is conceptually similar to prior work,
we accept the random access aspect of irregular applications and optimize for throughput rather than low latency.
Our DSM must support enough memory concurrency to tolerate the latency of the network.

@@

\paragraph{Multithreading} Grappa uses multithreading to tolerate memory
latency. This is a well known technique. Hardware implementations include the
Tera MTA~\cite{tera:mta1}, Cray XMT~\cite{feo:xmt}, Simultaneous
multithreading~\cite{tullsen:smt}, MIT Alewife~\cite{agarwal:alewife},
Cyclops~\cite{almasi:cyclops}, and even GPUs~\cite{gpus}. As we describe in
this paper, Grappa implements its own software-based multithreading with a
lightweight user-mode task scheduler to multiplex \emph{thousands\/} of tasks
on a single processing core. The large number of tasks is required to tolerate
the very high inter-node communication latency of commodity networks. Grappa's
task library employs several optimizations: an extremely fast task switch, a
small task size, and judicious use of software prefetch instructions to bring task state
into the cache sufficiently long before that task is actually scheduled. The main difference
between Grappa's support for lightweight threads and prior work such as
QThreads~\cite{Wheeler08qthreads:an} and
Capriccio~\cite{Behren03capriccio:scalable} is the use of prefetching by the scheduler, which is needed for good performance
when multiplexing such a large number of tasks.

>>>>

\paragraph{Distributed graph processing} While \Grappa is a general runtime
system for any large-scale concurrent application, one of its killer
applications is graph analysis. Therefore, distributed graph processing
frameworks like Pregel~\cite{pregel:2010} and Distributed
GraphLab~\cite{distgraphlab:vldb12} are related. Pregel adopts a
bulk-synchronous parallel (BSP) execution model, which makes it inefficient on
workloads that could prioritize vertices. GraphLab overcomes this limitation
with an execution mode that schedules vertex computations individually,
allowing prioritization, which gives faster convergence in a variety of
iterative algorithms. GraphLab, however, imposes a rigid computation model
where programmers must express computation as transformations on a vertex and
its edge list only, with information only from adjacent vertexes. Pregel is
only slightly less restrictive, as the input data can be any vertex in the
graph. \Grappa also supports dynamic parallelism with asynchronous execution,
but parallelism is expressed as tasks or loop iterations, which is a far more
general programming model for irregular computation tasks.
PowerGraph~\cite{powergraph:osdi12} improves the performance of GraphLab for
real-world graphs with power-law degree distributions by using a vertex cut
for graph partitioning. If the size of the partial result of applying a subset
of edge data to the vertex is constant, then there is a good bound on the
communication required to keep vertex replicas coherent. We do not explore
these algorithmic techniques in this work, but they are complimentary and
would improve the performance of graph applications on \Grappa.

\paragraph{Software distributed shared memory}
% Grappa supports a global memory namespace, presenting a programming
% abstraction similar to that of previous software DSM systems but with
% differences attuned to irregular application demands.
Many traditional software DSM systems are page based~\cite{Treadmarks,munin}
and aim to hide the fact that they are built in software from applications by
exploiting the processor's paging mechanisms, therefore relying heavily on
locality. Instead, \Grappa, like partitioned global address space (PGAS)
models, implements its DSM at the language, rather than the system level.
Languages such as Chapel~\cite{Chamberlain:2007}, X10~\cite{X10:2005}, and
UPC~\cite{upc:2005} make accesses to shared structures look like normal
references. As we describe later, \Grappa chooses a middle ground, where global
addresses are explicit in the API and local accesses are emitted
conventionally by the compiler.

Past DSM work was mostly page-based, so it could directly take advantage of a
network's RDMA support to move large page-sized blocks of data from node to
node \TODO{citation}. However, when commodity networks move small blocks of
data (a few bytes), only a fraction of the available bandwidth is achieved. \Grappa had to include significant aggregation support to make effective use of RDMA.

In summary, while \Grappa's DSM system is conceptually similar to prior work,
we accept the random access aspect of irregular applications and optimize for
throughput rather than low latency. Our DSM system must support enough memory
concurrency to tolerate the latency of the network.

@@

\paragraph{Multithreading} \Grappa uses multithreading to tolerate memory
latency. This is a well known technique. Hardware implementations include the
Tera MTA~\cite{tera:mta1}, Cray XMT~\cite{feo:xmt}, Simultaneous
multithreading~\cite{tullsen:smt}, MIT Alewife~\cite{agarwal:alewife},
Cyclops~\cite{almasi:cyclops}, and even GPUs~\cite{gpus}. As we describe in
this paper, \Grappa implements its own software-based multithreading with a
lightweight user-mode task scheduler to multiplex \emph{thousands\/} of tasks
on a single processing core. The large number of tasks is required to tolerate
the very high inter-node communication latency of commodity networks. \Grappa's
task library employs several optimizations: an extremely fast task switch, a
small task size, and judicious use of software prefetch instructions to bring task state
into the cache sufficiently long before that task is actually scheduled. The main difference
between \Grappa's support for lightweight threads and prior work such as
QThreads~\cite{Wheeler08qthreads:an} and
Capriccio~\cite{Behren03capriccio:scalable} is the use of prefetching by the scheduler, which is needed for good performance
when multiplexing such a large number of tasks.
